Normalizations are not what you think; Explicit Scale Simulation in ALDEx2

### 2024-07-22 go3.Rmd
focussed on dispersion version

---
title: "Normalizations are not what you think; Explicit Scale Simulation in ALDEx2
"
shorttitle: "Scale ALDEx2"
author:
  - name: Greg Gloor
    affiliation: Department of Biochemistry, University of Western Ontario
    email: ggloor@uwo.ca
  - name: Michelle Pistner Nixon
    affiliation: College of Information Sciences and Technology, Pennsylvania	 State University
  - name: Justin Silverman
    affiliation: College of Information Sciences and Technology and Department of Medicine, Pennsylvania	 State University
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
output:
  nucleic-acids-research.csl: pdf_document
package: ALDEx2
abstract: In many high-throughput sequencing (HTS) studies, sample-to-sample variation  in sequencing depth is  not driven by  variation in the scale (e.g., total size, microbial load, or total gene  expression) of the underlying biological systems being measured but rather is driven by technical factors. Typically the technical variation  is addressed using some form of statistical normalization.  Normalizations are data or parameter transformations that remove unwanted  technical variation in the hopes of facilitating analyses sensitive to scale;  e.g., differential abundance and differential expression analyses. Recently we  showed that any normalization makes implicit assumptions about the unmeasured  system scale and that errors in these assumptions can lead to dramatic increases in  false positive and false negative rates. Here we describe updates to the ALDEx2 R  package that mitigate these problems by directly modeling uncertainty in the  unmeasured system scale through the use of a \textit{scale model}. Scale  models generalize the idea of normalizations and can be thought of as  explicitly modeling potential error in the chosen normalization. Beyond  enhancing the robustness of HTS analyses, the use of scale models within ALDEx2   enhances the transparency and reproducibility of analyses by making implicit  normalizing assumptions an explicit part of the model building process.
---

@@

High-throughput sequencing (HTS) is a ubiquitous tool used to explore many biological phenomenon such as gene expression (single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial community composition (16S rRNA gene sequencing, shotgun metagenomics) and differential enzyme activity (selex, CRISPR killing). HTS proceeds by taking a sample from the environment, making a library, multiplexing (merging) multiple libraries together, and then applying a sample of the multiplexed library to the flow cell. Each of these steps can be though of as a sampling step: a pool of DNA that could be measured is sampled and only that subsampled DNA is carried over to subsequent steps. Within each sampling step, the connection between the size of the sampled DNA pool and the scale (e.g., size, microbial load, or total gene expression) of the measured biological system is degraded or lost. Increasingly, researchers are turning to modified experimental protocols (e.g., DNA spike-ins or cell counting) in an attempt to recover the lost biological variation in scale; see for example [@Vandeputte:2017aa;@Props:2017aa]. However, these protocols often do not recover meaningful biological variation but instead recover scale variation at some intermediate step in the sample preparation protocol (e.g., at the stage where the DNA spike-in was added). In short, the disconnect between sample-to-sample variation in sequencing depth and biological variation in scale remains an outstanding challenge.

Biological variation in scale can represent an important unmeasured confounder in HTS analyses [@Lovell:2015]. For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than  non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis [@Loven:2012aa]. In addition, wild-type and mutant strains of cell lines, yeast or bacteria often have different growth rates, which would affect our ability to identify truly differentially abundant genes [@Yoshikawa:2011aa]. As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance between the healthy and bacterial vaginosis states [@Zozaya:2010], and the composition between these states is dramatically different [@Ravel:2010;@Hummelen:2010]. Thus, a full description of these systems includes both relative change (composition) and absolute abundance (scale), yet current methods access only the compositional information.

To be more concrete, let \(\mathbf{Y}\) denote an \(D \times N\) matrix of sequence counts with elements \(\mathbf{Y}_{dn}\) denoting the number of measured DNA molecules mapping to feature \(d\) (e.g., a taxon or gene) in sample \(n\). Let \(\mathbf{W}\) also represent a \(D \times N\) matrix but with elements \(\mathbf{W}_{dn}\) representing the true (as opposed to measured) amount of class \(d\) in the biological system from which sample \(n\) was obtained. Succinctly, \(\mathbf{Y}\) denotes the observed HTS data whereas \(\mathbf{W}\) denotes the true counts in the system under study. We can think of \(\mathbf{W}\) as consisting of two parts, the scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition \(\mathbf{W}^{\parallel}\) (i.e., proportions). That is, \(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements \(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while \(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements \(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\). Note that with these definitions \(\mathbf{W}\) can be written as the element-wise combination of scale and composition: \(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\). The technical variation in sequencing depth (\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that observed data \(\mathbf{Y}\) provides us with information about the system composition \(\mathbf{W}^{\parallel}\) but little to no information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et al. 2011).

Early on, researchers recognized the potential confounding effects of the non-biological variation in sequencing depth and approached the problem using various forms of \textit{normalization}. The idea of normalization is to remove this variation thus making the samples commensurate and to correct any minor asymmetries in the data. A large number of normalizations were developed. Rarefaction [@Hughes:2005tu] was used in the microbiome field, while transcriptome investigators developed and used proportions and derivatives (reads per kilobase per million, and transcripts per kilobase per million) [@Mortazavi:2008; @wagner:tpm], the relative log expression (RLE) [@Anders:2010] and trimmed mean of M values (TMM) [@Robinson:2010a]. The Centered Log Ratio (CLR) [@aitchison1982]  was proposed as a unifying normalization [@fernandes:2014]. With the exception of rarefaction, all of these normalizations are data transformations that can be stated as ratios of the form \(\hat{\mathbf{Y}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\); the major differences being how the denominator is chosen and whether the ratio is subsequently always log-transformed or not.

Recently, Nixon et al. [-@nixon2023scale] suggested that the challenge of non-biological variation in sequencing depth be viewed as a problem of partially-identified models. Through this lens, they showed that \emph{all} normalizations make some assumption about the system scale \(\mathbf{W}^{\perp}\) in order to identify the chosen model. Moreover, these assumptions are often implicit, difficult to interpret, and provided different outputs when applied to the same dataset [@Dillies:2013;@Weiss:2017aa]. Intuitively, normalizations in widespread use assume that either one sample can be chosen as a reference to which the others are scaled (TMM), or that different sub-parts of each sample maintain a constant scale across samples (RLE, CLR and derivatives). Further, Nixon et al. [-@nixon2023scale] showed  that the original ALDEx2 model made a strict assumption about scale through the CLR normalization. They showed that this assumption could be remedied since ALDEx2 permits scale to be easily incorporated into its framework. In this paper, we discuss these modifications and show how including scale uncertainty can greatly improve modeling and lead to more robust and reproducible results.

@@

The ALDEx2 R package [@fernandes:2013] is a general purpose toolbox for Bayesian modeling of HTS data. For brevity and concreteness, we discuss ALDEx2 in its simplest form as a tool for estimating Log-Fold-Changes (LFC) (e.g., differential abundance or differential expression analysis), but note that it can be used to fit more complex linear models. At a high-level, ALDEx2 involves three steps: 1) a Bayesian model is used to estimate \(\mathbf{W}^{\parallel}\) given observations \(\mathbf{Y}\); 2) a normalization is used to estimate \(\mathbf{W}\) given the estimate \(\mathbf{W}^{\parallel}\); 3) the \(\mathbf{W}\) estimates are used to estimate LFC as part of differential abundance (or expression) analyses. For more details on ALDEx2 see Fernandes et al. [-@fernandes:2013].

By default, ALDEx2 uses the CLR normalization which can be written as \(\log \mathbf{W}_{dn}=\log [\mathbf{W}^{\parallel}_{dn}/G_{n}]\) where \(G_{n}\) denotes the geometric mean of the vector \((W^{\parallel}_{1n}, \dots, W^{\parallel}_{Dn})\) (see Supplemental Materials for how the CLR connects to Shannon entropy). Nixon et al. (2023) showed that in this context the CLR normalization equates to an implicit assumption that \(\mathbf{W}^{\perp}_{n}=1/G_{n}\). They showed that even slight errors in this assumption introduced bias into LFC estimate that was not accounted for in estimates of uncertainty (e.g., confidence intervals or credible sets). In fact, they showed that the only way in which the ALDEx2 model, or any model that uses  normalizations, could ever be calibrated (e.g., control Type-I Error rates) was if this assumption was exactly true. While this assumption may be approximately correct in some instances it is surely never exactly true. In support of this assertion it is known that false discovery rates vary widely by analysis and normalization method [@hawinkel2017;@Li:2022aa]. As a result, Nixon et al. (2023) suggested that models should consider potential error in the assumptions implied by normalizations.

Nixon et al. (2023) proposed the concept of a \textit{scale model} as a generalization of normalizations that account for potential error. They showed how scale models can be incorporated into ALDEx2, turning the ALDEx2 model into a specialized type of statistical model which they called a \textit{Scale Simulation Random   Variable} (SSRV). Recall that \(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).  As the HTS data only informs directly on $\mathbf{W}^{\parallel}_{dn}$, they proposed to augment analyses with a model for $\mathbf{W}^{\perp}_{n}$ (the scale model). Since the CLR normalization makes the assumption \(\mathbf{W}^{\perp}_{n}=1/G_{n}\), the CLR normalization can be generalized by considering probability models for the scale \(\mathbf{W}^{\perp}_{n}\) that have mean \(1/G_{n}\). For example, the following scale model generalizes the CLR:

\[\log \mathbf{W}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(0, \gamma^{2})\]

where \(\gamma\) is a tunable parameter that controls the degree of uncertainty in the CLR assumption and \(x_{n}\) denotes a binary condition indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control). Beyond theoretical studies, they demonstrated through the analysis of simulated and real data the remarkable performance improvements that could result from incorporating this type of scale uncertainty into analyses. Subsequently, we have made those modifications a permanent fixture of ALDEx2 which now represents the first software package designed for SSRV-based inference.

Beyond concerns of fidelity and rigor, scale models also enhance the reproducibility and transparency of HTS analyses. This occurs because rather than using an implicit assumption (e.g., \(\log \mathbf{W}^{\perp}=1/G_{n}\) for the CLR normalization), scale models make this an explicit part of the model building process. While it is beyond the scope of the present article, we note that there are many ways of building scale models that enhance the interpretability of the parameters and assumptions. Moreover, the scale model listed above, which is now the default in ALDEx2, also has an appealing interpretation. A detailed description of these points as well as the mathematical underpinnings of the updated ALDEx2 R package is forthcoming and will be made available as a companion publication. Here, we simply treat the parameter \(\gamma\) in the default scale model as an abstract term that represents uncertainty in the CLR assumption and note that larger values of $\gamma$ correspond to more uncertainty in this assumption. We also show that the ratio of scales between conditions can be used to build a full scale model when the conditions have dramatically different underlying scales.

An advantage of incorporating scale is that analyses can be made much more robust such that actual or potential differences in scale can be tested and accounted for explicitly. The examples below and in the supplement show how incorporating scale provides robust and interpretable differential abundance estimates in several different datasets.

@@

Gierliński et al. [-@Gierlinski:2015aa] conducted a highly replicated yeast transcriptome experiment comparing a wild-type strain with a snf1 gene knockout. This dataset has been used to argue that only a small number of replicates are needed to identity differentially abundant genes and makes the recommendation that some tools should be used in some conditions because each tool has different intrinsic statistical power and Type 1 and Type 2 errors [@Schurch:2016aa]]. This guidance runs counter to standard statistical practice where power is intrinsically linked to sample size [@Halsey:2015aa], yet the concept of sample-size independent power is entrenched in all fields that use HTS as an experimental readout. Through the lens of this paper, this is a hallmark of unacknowledged bias: a false certainty in the scale of the data brings about an increasing false precision with increasing sample sizes. Furthermore, the yeast transcriptome experiment has an unacknowledged difference in scale between conditions: yeast deficient for snf1 are smaller, grow more slowly and are sensitive to a variety of common agents that cause cell stress [@Yoshikawa:2011aa].  Thus, by adding scale uncertainty, we can be more confident that the results of the analysis reflect true biological differences and not differences due to the assumptions made by the normalization method.

We start by examining the dispersion (variance) of the data as counts and as the logarithm of the normalized counts when calculated by DESeq2 (log(RLE)) or by ALDEx2 (CLR). The raw data counts derived from sequencing are overdispersed with the mean value being less than the variance [@Robinson:2010a;@Robinson:2010] as seen in Panel A of Figure 1. However, the actual analysis of differential abundance is performed on the logarithm of the normalized counts [@Robinson:2010;@Love:2014aa;@Andrews:2019aa], which includes the CLR [@fernandes:2013] yet the mean-dispersion distribution of these log-ratio transformed data is quite different as shown in panels B and C of Figure 1 and as noted elsewhere [@fernandes:2013;@Love:2014aa].

```{r, echo=F, warning=F, message=F,comment=F, result=F} 
library(DESeq2)
devtools::load_all('~/Documents/0_git/ALDEx_bioc')
#library(ALDEx2)
```
```{r dispersion, echo=F, warning=F, message=F,comment=F, result=F, cache=T, fig.cap="Plot of abundance v dispersion for a typical transcriptome dataset as counts,	 as logarithms of counts, and as clr values. Panel A shows that the data are over-dispersed relative to a Poisson distribution which is represented by the dashed line when plotted on a log-log scale. Panel B shows that the relationship between the mean and the dispersion calculated in DESeq2, here the standard error (SE) of the mean, is very different when the data are log-transformed first. Panel C shows the equivalent values calculated by ALDEx2	 in which the expected clr value for each transcipt are plotted vs. the expected dispersion. The red line in each panel shows the loess line of fit to the mid-point of the distributions. In panels B and C the amount of dispersion reaches a minimum at moderate	 values.  The dashed orange line in panel A is the line of equivalence, and in panel B and C is the minimum y value. The values below the dashed grey line in panels B and C represent those below the first decile of dispersion." }

url <- "https://raw.githubusercontent.com/ggloor/datasets/main/transcriptome.tsv"
yst <- read.table(url, header=T, row.names=1)
# remove the one gene with 0 reads

yst <- yst[rownames(yst) != "YOR072W-B",]

# Gierlinski:2015aa
yst[,c('SNF2.6', 'SNF2.13','SNF2.25','SNF2.35')] <- NULL 
yst[,c('WT.21','WT.22','WT.25','WT.28','WT.34','WT.36')] <- NULL

conds <- c(rep('S', 44), rep('W', 42))
coldata <- data.frame(conds)

# DESeq2
load('analysis/res.Rda')

#ALDEx2
load('analysis/x.all.Rda')
load('analysis/x.s.all.Rda')

# with gamma = 1
load('analysis/x.s.all.Rda')
sig.des <- which(res@listData$padj < 0.01)
sig.ald <- which(x.all$we.eBH < 0.01)
sig.s.ald <- which(x.s.all$we.eBH < 0.01)

# get and plot mean/var or mean/SE for counts
# and log counts
yst.mn <- apply(yst, 1, mean)
yst.mn.log <- apply(yst, 1, function(x) mean(log2(x)))
yst.mn.log[is.infinite(yst.mn.log)] <- 0

yst.v <- apply(yst, 1, var)

par(mfrow=c(1,3))
plot(log2(yst.mn), log2(yst.v), pch=19, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='Mean count', ylab='Variance')
#points(log2(yst.mn[sig.des]), log2(yst.v[sig.des]), pch=19, cex=0.5, col='orange')
def <- data.frame(log2(yst.mn),log2(yst.v))
lines(lowess(def, f=0.1), col=2, lwd=2)
title('A: M v. V', adj=0, line= 0.8)
abline(0,1, col=rgb(1,0.6,0,0.5), lwd=2, lty=2)

l.df <- data.frame(yst.mn.log, res@listData$lfcSE)
plot(l.df, pch=19, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='Mean log2(count)', ylab='SE log2(count)',ylim=c(0,0.4))
lines(lowess(l.df, f=0.25), col=2, lwd=2)
abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
 abline(h=0.0275, col='grey', lty=2)
title('B: log2(M) v. log2(V)', adj=0, line= 0.8)

l.df <- data.frame(x.all$rab.all, x.all$diff.win)
plot(x.all$rab.all, x.all$diff.win, cex=0.5, col=rgb(0,0,0,0.2),
  xlab='E(clr)', ylab='E(disp)')
#points(x.all$rab.all[sig.ald], x.all$diff.win[sig.ald], col='orange', cex=0.4)
lines(lowess(l.df, f=.1), col=2, lwd=2)
abline(h=min(lowess(l.df)$y), col=rgb(1,0.6,0,0.5), lwd=2, lty=2)
 abline(h=0.208, col='grey', lty=2)
title('C: E(clr) vs. E(dispersion)', adj=0, line= 0.8)
```

>>>>

---
title: "Normalizations are not what you think; Explicit Scale Simulation in ALDEx2
"
shorttitle: "Scale ALDEx2"
author:
  - name: Greg Gloor
    affiliation: Department of Biochemistry, University of Western Ontario
    email: ggloor@uwo.ca
  - name: Michelle Pistner Nixon
    affiliation: College of Information Sciences and Technology, Pennsylvania	 State University
  - name: Justin Silverman
    affiliation: College of Information Sciences and Technology and Department of Medicine, Pennsylvania	 State University
citeproc: TRUE
cite-method: citeproc
bibliography: /Users/ggloor/Library/texmf/bibtex/bib/bibdesk_refs.bib
csl: /Users/ggloor/Documents/0_git/csl_styles/nucleic-acids-research.csl
pdf-engine: latex
output:
  pdf_document 
header-includes:
  - \usepackage{setspace}
package: ALDEx2
abstract: In  high-throughput sequencing (HTS) studies, sample-to-sample variation in sequencing depth is driven by technical factors, and not by variation in the scale (e.g., total size, microbial load, or total gene expression) of the underlying biological systems. Typically a statistical normalization is used  to remove unwanted technical variation in the data or the parameters of the model to enable analyses that are sensitive to scale; e.g., differential abundance and differential expression analyses. Recently we showed that all normalizations make implicit assumptions about the unmeasured system scale and that errors in these assumptions can lead to dramatic increases in false positive and false negative rates. Here we describe updates to the ALDEx2 R package that mitigate these problems by directly modeling uncertainty in the unmeasured system scale through the use of a \textit{scale model}. Scale models generalize the idea of normalizations and can be thought of as explicitly modeling the error in normalization by examining a distribution over all possible  normalizations.  Beyond enhancing the robustness of HTS analyses,  the use of scale models within ALDEx2 enhances the transparency and  reproducibility of analyses by making implicit normalizing assumptions an explicit part of the model building process.

---

@@

High-throughput sequencing (HTS) is a ubiquitous tool used to explore
many biological phenomenon such as gene expression (single-cell
sequencing, RNA-sequencing, meta-transcriptomics), microbial community
composition (16S rRNA gene sequencing, shotgun metagenomics) and
differential enzyme activity (selex, CRISPR killing). HTS proceeds by
taking a sample from the environment, making a library, multiplexing
(merging) multiple libraries together, and then applying a sample of the
multiplexed library to the flow cell. Each of these steps is a 
compositional sampling step as only a fixed-size subsample of nucleic acid is 
carried over to subsequent steps. Thus, with
each sampling step the connection between the size of the sampled DNA
pool and the scale (e.g., size, microbial load, or total gene
expression) of the measured biological system is degraded or lost.
Increasingly, researchers are turning to modified experimental protocols
(e.g., DNA spike-ins or cell counting) in an attempt to recover the lost
biological variation in scale; see for example  [@Vandeputte:2017aa;@Props:2017aa]. 
However, these protocols often do not recover
meaningful biological variation but instead recover scale variation at
the intermediate step in the sample preparation protocol where the DNA spike-in was added. 
In short, the disconnect
between sample-to-sample variation in sequencing depth and biological
variation in scale remains an outstanding challenge.

The analysis of  HTS data suffers from several known problems that can be
traced, in whole or in part, to  misspecification of scale. The first issue 
is poor control of the false discovery rate (FDR) [@Thorsen:2016aa;@Quinn:2018aa], 
  exhibited as dataset-dependent 
FDR control.  The FDR problem is connected to the double-filtering approach
that is often used, but which is known not to have appropriate FDR control 
[@Zhang:2009aa;@Ebrahimpoor:2021aa]. The second issue is poor performance when 
analyzing asymmetric data; data that arises because of a bias in the direction of change.
This type of data frequently arises in in-vitro selection experiments (SELEX), 
transcriptome analysis, and microbiome analysis [@Wu2021]. The third issue is that these
problems become more pronounced as more samples are collected; that is,
more information results in a worsening of the accuracy of the analysis [@nixon2023scale,@Nixon2024B]. 
The final problem is that these data are relative data, i.e. compositional, 
and substantial effort has been put into removing this constraint.

The first three problems were recently shown by Nixon and Silverman [@nixon2023scale] to be a result of a mismatch between the underlying size or scale of the system and the assumptions of the normalizations used for the analysis of HTS.  Biological variation in scale often represents an important unmeasured confounder in HTS analyses [@Lovell:2015]. For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than  non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis [@Loven:2012aa]. In addition, wild-type and mutant strains of cell lines, yeast or bacteria  have different growth rates and RNA contents under different conditions, which  affect our ability to identify truly differentially abundant genes [@Scott:2010;@Yoshikawa:2011aa;@Lin:2018aa]. As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance between the healthy and bacterial vaginosis states [@Zozaya:2010], and the composition between these states is dramatically different [@Ravel:2010;@Hummelen:2010]. Thus, a full description of any of these systems includes both relative change (composition) and absolute abundance (scale).  Current methods access only the compositional information yet make implicit assumptions about the scale.

Recently, Nixon et al. [-@nixon2023scale] showed that the challenge of non-biological variation in sequencing depth be viewed as a problem of partially-identified models. They showed that \emph{all} normalizations make some assumption about  scale. These implicit assumptions are often  difficult to interpret, and different normalizations provide different outputs when applied to the same dataset [@Bullard:2010;@Dillies:2013;@Thorsen:2016aa;@Weiss:2017aa]. Intuitively, normalizations in widespread use  assume that either all samples have the same scale, e.g. proportions, rarefaction [@Hughes:2005tu], RPKM [@Mortazavi:2008; @wagner:tpm], etc; or that a subset of features in one sample can be chosen as a reference to which the others are scaled e.g. the TMM,[@Robinson:2010a], or LVHA [@Wu2021]; or that different sub-parts of each sample maintain a constant scale across samples e.g. the RLE [@Anders:2010];  or that the geometric mean of the parts is appropriate e.g the CLR [@aitchison1982] and its derivatives.

The original ALDEx2 [@fernandes:2013] model made a strict assumption about scale through the CLR normalization [@nixon2023scale]. While this assumption could be useful in many cases it could  never be exactly true, and others have shown that it is not always the best option [@Yerke:2024aa]. Nixon et al. [-@nixon2023scale] showed that better scale assumptions resulted in more reproducible data analysis including better control of both false positive and false negative results. In essence ALDEx2 has been modified to explicitly model the scale over a range of reasonable normalization parameters. Here, we briefly introduce these modifications and then show how scale uncertainty can greatly improve modeling in transcriptome and meta-transcriptome datasets to provide more robust and reproducible results.

# Implementation

To be  concrete, we let \(\mathbf{Y}\) denote the \emph{measured} \(D \times N\) matrix of sequence counts with elements \(\mathbf{Y}_{dn}\) indicating the number of measured DNA molecules mapping to feature \(d\) (e.g., a taxon or gene) in sample \(n\).  Likewise, we can denote \(\mathbf{W}\) as the \emph{true}  amount of class \(d\) in the biological system from which sample \(n\) was obtained. We can think of \(\mathbf{W}\) as consisting of two parts, the scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition \(\mathbf{W}^{\parallel}\) (i.e., proportions). That is, \(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements \(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while \(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements \(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\). Note that with these definitions \(\mathbf{W}\) can be written as the element-wise combination of scale and composition: \(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\).

All  the normalizations in current use are data transformations that can be stated as ratios of the form \(\hat{{\mathbf{W}}}_{dn}=\mathbf{Y}_{dn}/f(\mathbf{Y})\), where the denominator is determined by some function of the observation and the \(\hat{{\mathbf{W}}}\) notation indicates that the normalization is attempting to provide an estimate of the true value.  The technical variation in sequencing depth (\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that observed data \(\mathbf{Y}\) provides us with information about the system composition \(\mathbf{W}^{\parallel}\) but little to no information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et al. 2011).

@@

The ALDEx2 R package [@fernandes:2013] is a general purpose toolbox for Bayesian modeling of HTS data. For brevity, we discuss ALDEx2 in its simplest form as a tool for estimating the magnitude and statistical significance of Log-Fold-Changes (LFC) (e.g., differential abundance or differential expression analysis), but note that it can be used to fit more complex linear models. At a high-level, ALDEx2 involves three steps: 1) a Bayesian model is used to estimate \(\mathbf{\hat{W}}^{\parallel}\) given observations \(\mathbf{Y}\); 2) a normalization is used to estimate \(\mathbf{\hat{W}}\) given the estimate \(\mathbf{\hat{W}}^{\parallel}\); 3) the \(\mathbf{\hat{W}}\) estimates are used to estimate LFC as part of differential abundance (or expression) analyses. For more details on ALDEx2 see [@fernandes:2013;@nixon2023scale].

By default, ALDEx2 uses the CLR normalization which can be written as \(\log \mathbf{\hat{W}}_{dn}=\log (\mathbf{\hat{W}}^{\parallel}_{dn}/G_{n})\) where \(G_{n}\) denotes the geometric mean of the vector \((\hat{W}^{\parallel}_{1n}, \dots, \hat{W}^{\parallel}_{Dn})\). The CLR normalization equates to an implicit assumption that \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\)[@nixon2023scale,Nixon2024B] and even slight errors in this assumption introduces bias into the LFC estimate that is not accounted for when estimating uncertainty (e.g., confidence intervals or credible sets). In fact, Nixon et al. [-@nixon2023scale] showed that the only way in which the ALDEx2 model, or any normalization-based model, could ever be calibrated (e.g., control Type-I Error rates) was if this assumption was exactly true.  In support of this assertion it is known that false discovery rates vary widely by analysis, dataset, and normalization method [@hawinkel2017;@Li:2022aa,Nixon2024B]. To fix this, Nixon et al. [-@nixon2023scale] showed that models should incorporate potential error in the assumptions implied by normalizations.

Nixon et al. (2023) generalized the concept of normalizations by introducing  the concept of a \textit{scale model} to account for potential error. Scale models can be incorporated into ALDEx2, turning the ALDEx2 model into a specialized type of statistical model which they called a \textit{Scale Simulation Random   Variable} (SSRV). They did this by including a model for \(\mathbf{\hat{W}}^{\perp}_{n}\). Since the CLR normalization makes the assumption \(\mathbf{\hat{W}}^{\perp}_{n}=1/G_{n}\), the CLR normalization can be generalized by considering probability models for the scale \(\mathbf{\hat{W}}^{\perp}_{n}\) that have mean \(1/G_{n}\). For example, the following scale model generalizes the CLR:

\[\log \mathbf{\hat{W}}^{\perp}_{n} = -\log G_{n} + \Lambda x_{n} \qquad \Lambda \sim N(0, \gamma^{2})\]

where \(\gamma\) is a tunable parameter drawn from a Gaussian distribution that controls the degree of uncertainty in the CLR assumption and \(x_{n}\) denotes a binary condition indicator (e.g., \(x_{n}=1\) denotes case and \(x_{n}=0\) denotes control). We have made those modifications a permanent fixture of ALDEx2 which now represents the first software package designed for SSRV-based inference.

@@

It is standard practice in many fields of HTS, but especially transcriptomics, to use the dual cutoff approach graphically exemplified by volcano plots [@Cui:2003aa;@Schurch:2016aa] because not all statistically significant differences are biologically relevant. Paraphrasing this we can say that in in some types of datasets more samples leads to a majority of features being statistically significant. A detailed reasoning for this using modelled data is explained in two reports by Nixon et al. [-@nixon2023scale;-@Nixon2024B], which shows that scaled models completely address this analytic issue.

We use the data of Gierliński et al. [-@Gierlinski:2015aa] who conducted a highly replicated yeast transcriptome experiment comparing a wild-type strain with a snf2 gene knockout, \(\Delta\)snf2. This dataset has been used to argue that a dual cutoff approach is appropriate to limit the number of significant parts and for purposes of reproducibility [@Schurch:2016aa]. However, in this benchmarking study,  the number of significantly different transcripts varied between 65% to >80% of all transcripts depending on the tool; in essence the desired behavior was that almost all transcripts were significantly different. The guidance on reproducibility runs counter to standard statistical practice where power is intrinsically linked to sample size and very large sample sizes are indeed desired [@Halsey:2015aa]. Furthermore, the dual-cutoff approach is  known not to provide appropriate FDR control [@Zhang:2009aa; @Ebrahimpoor:2021aa].  Schurch et al. -@Schurch:2016aa further suggested that different tools might be better in some conditions or datasets than others because each tool has different intrinsic statistical power and Type 1 and Type 2 errors. Through the lens of scale uncertainty, the behaviour of the tools in this study show unacknowledged bias; false confidence in the precision of the estimate as sample size increases because o mis-specification  in the scale of the data. This certainty is driven by the assumptions of the tool not the actual experiment being investigated [@Nixon2024B].

```{r yst-res, echo=F, warning=F, message=F,comment=F,}
#ALDEx2
library(ALDEx2)
load(file="analysis/yst.all.Rda")
load(file="analysis/yst.s.all.Rda")
load(file="analysis/yst.1.all.Rda")
# DESeq2
load('analysis/res.Rda')

# with gamma = 0.5
load('analysis/x.s.all.Rda')
sig.des <- which(res@listData$padj < 0.05)
sig.ald <- which(yst.all$we.eBH < 0.05)
sig.s.ald <- which(yst.s.all$we.eBH < 0.05)

sig.all <- yst.all$we.eBH < 0.05
sig.t <- yst.all$we.eBH < 0.05 & abs(yst.all$diff.btw) > 1.4
sig.s <- yst.s.all$we.eBH < 0.05
sig.1 <- yst.1.all$we.eBH < 0.05
```

Using either DESeq2 or ALDEx2, a majority of transcripts are statistically significantly different between groups  with a Benjamini-Hochberg [@benjamini:1995] false discovery rate (FDR) of 0.05; i.e. `r length(sig.des)` (`r round(length(sig.des)/nrow(yst.all), 2)*100`%, DESeq2) or `r length(sig.ald)` (`r round(length(sig.ald)/nrow(yst.all), 2)*100`%, ALDEx2) of the `r nrow(yst.all)` transcripts. Such  large numbers of significant transcripts seems biologically unrealistic  and furthermore  breaks the necessary assumption made by the normalization methods that the majority of the features must be invariant. That `r length(setdiff(sig.ald, sig.des))` transcripts are identified by ALDEx2 and not DESeq2, while DESeq2 identifies `r length(setdiff(sig.des, sig.ald))` transcripts that ALDEx2 does not, suggests that the choice of normalization plays a role in which results are returned as significant and that some if not the majority are driven by technical differences in the analysis [@Soneson:2013;@maza2013;@Dillies:2013;@Weiss:2017aa; @Schurch:2016aa].

```{r plot1, echo=F, fig.dim=c(5,6), fig.cap="Effect and volcano plots for unscaled and scaled transcriptome analysis. ALDEx2 was used to conduct a differential abundance (DA) analysis on the yeast transcriptome dataset. The results were plotted to show the relationship between difference and dispersion using effect plots ) or difference and the Benjamini-Hochberg corrected p-values (volcano plot). Panels A,C are for the unscaled analysis, and Panels B,D are for the scaled analysis. Each point represents the values for one transcript, with the color indicating if that transcript was significant in the scaled analysis and unscaled analysis (red) or in the unscaled analysis only (orange). Points in grey are not statistically signficantly different with any analysis. The horizontal dashed lines represent a log2(difference) of \\(\\pm 1.4\\)."}

par(mfrow=c(2,2))

#plot(res@listData$lfcSE*sqrt(ncol(yst)), res@listData$log2FoldChange, xlim=c(0,5), 
#  col=rgb(0,0,0,0.1), xlab='LFC SD', ylab='log2 Difference')
#title('A: DESeq2 effect', adj=0, line= 0.8)
#points(res@listData$lfcSE[sig.des]*sqrt(ncol(yst)),
#  res@listData$log2FoldChange[sig.des], col=rgb(1,.66,0,0.5), 
#  pch=19, cex=0.5)
#points(res@listData$lfcSE[sig.s.ald]*sqrt(ncol(yst)),
#  res@listData$log2FoldChange[sig.s.ald], col=rgb(1,0,0,0.5), 
#  pch=19, cex=0.5)
#abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')

plot(yst.all$diff.win, yst.all$diff.btw, col=rgb(0,0,0,0.1), 
   xlab='Dispersion', ylab='log2 Difference')
title(expression(paste("A: Effect: ", gamma, " = 0")), adj=0, line= 0.8)
points(yst.all$diff.win[sig.ald], yst.all$diff.btw[sig.ald], 
  col=rgb(1,.66,0,0.3), cex=0.5, pch=19)
	abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')
points(yst.all$diff.win[sig.s.ald], yst.all$diff.btw[sig.s.ald], 
  col=rgb(1,0,0,0.6), cex=0.5, pch=19)
	abline(h=c(-1.4,1.4), lty=2,lwd=2, col='grey')
